Exploring dimensions of biodiversity in Japanese ferns

### 2021-05-27 ms/manuscript.Rmd
Parameterize markdown to render to doc and pdf separately

---
title: "Exploring dimensions of biodiversity in Japanese ferns"
bibliography: 
  - references.bib
  - references_other.yaml
mainfont: Roboto-Regular.ttf
fontsize: 11pt
mainfontoptions: 
- BoldFont=Roboto-Bold.ttf
- ItalicFont=Roboto-Italic.ttf
- BoldItalicFont=Roboto-BoldItalic.ttf
output: 
  bookdown::pdf_document2:
    citation_package: none # Use pandoc default, otherwise csl can't be used for reference formatting
    toc: no
    number_sections: no
    fig_caption: yes
    keep_tex: yes
    latex_engine: xelatex
    pandoc_args: [ "--csl", "journal-of-biogeography.csl"]
    includes:
      in_header: newphyt.sty
header-includes: 
  \usepackage{float}
  \makeatletter\renewcommand*{\fps@figure}{H}\makeatother
editor_options: 
  chunk_output_type: console
# nocite: | 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, results = "hide", cache = FALSE)
```

```{r load-data, cache = FALSE}
# NOTE: This Rmd is meant to be knit as part of the drake plan (R/plan.R).
# If using interactively, run `source("_targets.R")` first.

tar_load(c(
  "occ_point_data_ferns",
  "occ_point_data_ferns_unfiltered",
  "biodiv_ferns_spatial",
  "k_taxonomy",
  "k_phylogeny",
  "signif_cells_protected_area",
  "protected_areas",
  "japan_shp",
  "phy_sig_results",
  "binary_sig_results",
  "traits_summary",
  "fern_traits",
  "japan_pterido_tree",
  "japan_fern_tree",
  "ppgi",
  "occ_point_data_ferns",
  "comm_ferns_full",
  "comm_ferns",
  "japan_points_raw",
  # "plastome_calibration_dates",
  # FIXME: remove until phylo analysis is working "plastome_tree",
  "green_list"
))

# Load captions
source(here::here("ms/captions.R"))
```

@@

\clearpage

@@

We used traits originally compiled for identification of ferns and lycophytes of Japan as follows [@Ebihara2016b; @Ebihara2017]. Continuous traits were measured on 10 randomly selected specimens per species. Fertile (\ie, spore-bearing) and sterile fronds were measured separately, then averaged. Qualitative traits were scored by observing voucher specimens. All qualitative traits were scored in binary format; \eg, a single trait with three states was formatted as three binary traits. From the original trait list, we selected only traits with putative ecological function (`r table("traits-used")`), and removed any that had Pearson correlation coefficient >0.6 or fewer than three observations of a given trait state.

@@

We downloaded climate data at the 2.5 minute scale from the WorldClim database using the `getData` function in the R package `raster`, and extracted four variables related to our research questions as follows: mean annual temperature (bio1; hereafter "temperature"), temperature seasonality (bio4), annual precipitation (bio12; hereafter "precipitation"), and precipitation seasonality (bio15). We identified the intersection of the climate data at the 2.5 minute scale with the 0.2° grid-cells, then calculated the mean of each climatic variable for each 0.2° grid-cell. The R package 'sf' was used for GIS analyses.

@@

We tested for phylogenetic signal in continuous traits using two alternative metrics, Blomberg's *K* [@Blomberg2003] and Pagel's lambda [@Pagel1999]. *K* describes the ratio between the observed variance in a trait vs. the amount of variance expected under Brownian Motion (BM); lambda is a scaling parameter that transforms the phylogeny such that trait values fit those most closely expected under BM. Both lambda and *K* = 1 under BM; lower values indicate less signal (overdispersion) and higher values indicate more signal (clustering). We measured *K* and lambda using the \function{phylosig} function in the R package \package{phytools} [@Revell2012], which also calculates significance based on a randomization test for *K* and a likelihood ratio test for lambda. We tested for phylogenetic signal in qualitative (coded as binary) traits using Fritz and Purvis' *D* [@Fritz2010] with the \function{phylo.d} function in the R package \package{caper} [@Orme2018], which also assesses significance by comparing the observed value to a null distribution of random values and a simulated distribution of values expected under BM.

@@

We measured phylogenetic diversity (PD) using the Faith's PD (the total branch length connecting all terminal taxa in a community, including the root of the phylogeny) [@Faith1992] with the \function{PD} function in the R package \package{phyloregion}. To detect areas with unusually long or short branch lengths, we measured relative phylogenetic diversity (RPD), which is a ratio of observed PD to PD measured on a tree where all branch lengths have been set to equal length [@Mishler2014].

We measured functional diversity (FD) following the method of @Petchey2002. We first calculated functional distances using Gower's method [@Gower1971] on the trait data with the \function{gowdis} function in the R package \package{FD} [@Laliberte2010]. We weighted continuous and qualitative traits equally, weighted each binary component trait of each qualitative trait equally, and log-transformed and scaled continuous traits prior to calculating distances. We then used the distances to build a trait dendrogram using the \function{hclust} function in the \package{stats} R package [@RCoreTeam2020], and used the dendogram in place of a phylogenetic tree to calculate PD and RPD as described above. These functional analogs of PD and RPD are hereafter referred to as FD (functional diversity) and RFD (relative functional diversity).

We measured phylogenetic endemism (PE) [@Rosauer2009], which is the sum of all branches at a site inversely weighted by their range, using the \function{phylo{\_}endemism} function in the R package \package{phyloregion}.

@@

We analyzed bioregions using the framework of @Daru2020 and their R package \package{phyloregion}. Briefly, this involves first calculating beta diversity (change in composition between communities) using either species names (taxonomic bioregions) or phylogenetic distances (phylogenetic bioregions). For taxonomic bioregions, we calculated the turnover in species between sites due to species replacement [@Baselga2012] with the Sørensen index. For phylogenetic bioregions, we calculated phylogenetic dissimilarities based on the PhyloSor index [@Bryant2008]. The distances are used to construct a dendrogram, which is then split into into *k* bioregions. As the optimal number of bioregions cannot be known *a-priori*, we tested values of *k* from 1 to 20.

@@

Grid-cells with significantly low PD were almost all located on Honshu, the main island of Japan. A smaller number of cells had significantly high PD, and these were located mostly on smaller southern (Amami, Ryukyu, Izu, Ogasawara) islands or coastal areas (Izu Peninsula, Kii Peninsula). The area near Yakushima includes cells with both significantly high and low PD. Grid-cells with significantly low RPD were mostly observed in southern Honshu, particularly in and around Kyushu. Grid-cells with significantly high RPD are located both in the southern islands and northern Honshu and Hokkaido. Results of the randomization test for functional diversity were broadly similar to phylogenetic diversity, but with more significant cells for FD than PD and fewer significant cells for RFD than RPD.

@@

Nearly all grid-cells in islands south of Kyushu (Ryukyu and Ogasawara) have significant levels of phylogenetic endemism, with low (non-significant) endemism observed elsewhere. The vast majority of endemic grid-cells are mixed, with very few paleo- or neoendemic cells. Small clusters of significantly endemic cells were also observed northwest of Tokyo (Nagano prefecture), and in central Hokkaido (Daisetsuzan mountains). When we ran CANAPE with a reduced dataset including only taxa endemic to Japan to account for the border effect, there were fewer significant cells, but patterns were otherwise similar to the full dataset (`r s_figure("endemism-restricted")`).

@@

The extreme evolutionary distinctiveness of southern islands is both a major feature of the Japanese fern flora and a challenge for analysis. The tropical lineages occurring in these islands are likely responsible for the comparatively low (\ie, clustered) values of PD observed on Honshu (`r figure("rand-diversity")`A). If a more restricted null model was used that did not consider all taxa capable of dispersing anywhere within the study area, different results would likely be obtained for the randomization test [@Mishler2020]. However, ferns are known to have high dispersal abilities [@Tryon1970; @Barrington1993; @Moran2001]. In our dataset, the taxon with the greatest latitudinal distribution, `r taxon_with_widest_lat$text`, spans nearly the entire region (`r taxon_with_widest_lat$min_lat`°N to `r taxon_with_widest_lat$max_lat`°N), and `r n_widespread_taxa` taxa have latitudinal distributions spanning the Tokara strait in the south (29°N) to the Tsugaru strait in the north (41.5°N). Therefore, it seems reasonable to assume that ferns are capable of dispersing across the entire study area.

@@

\clearpage

<!-- Tables -->

```{r traits-used, results = "asis"}
trait_signif <- read_csv(here::here("data_raw/trait_functions.csv"))

traits_summary %>%
  left_join(trait_signif, by = "trait") %>%
  mutate(
    trait = case_when(
      trait == "frond_width" ~ "Frond width",
      trait == "number_pinna_pairs" ~ "No. pinna pairs",
      trait == "stipe_length" ~ "Stipe length",
      trait == "margin" ~ "Frond margin",
      trait == "shape" ~ "Frond shape",
      trait == "texture" ~ "Frond texture",
      TRUE ~ str_to_sentence(trait)
    ),
    trait_type = case_when(
      trait_type == "qualitative" ~ glue::glue("Qualitative (coded as {n} binary traits)"),
      trait_type == "continuous" & str_detect(trait, "pinna pairs") ~ glue::glue("Continuous (integer)"),
      trait_type == "continuous" ~ glue::glue("Continuous (cm)"),
      TRUE ~ str_to_sentence(trait_type)
    )
  ) %>%
  select(trait, Significance = significance) %>%
  column_to_rownames("trait") %>%
knitr::kable(format = "latex", booktabs = TRUE, escape = TRUE)

```

**`r table("traits-used")`**. `r table_cap("traits-used")`

\clearpage

<!-- Figures -->

@@

```{r japan-map, fig.width = 5, fig.height = 5}
# Print map of Japan with labels of relevant places

@@

# Make map
ggplot(japan) + 
  geom_sf() +
  geom_richtext(
    data = japan_points, 
    aes(x = lon_adj, y = lat_adj, label = label),
    size = 2.5,
    fill = NA, label.color = NA, # remove background and outline
    label.padding = grid::unit(rep(0, 4), "pt") # remove padding)
  ) +
  geom_point(
    data = filter(japan_points, query == "Tokyo"),
    aes(x = lon, y = lat)
  ) +
  map_theme()

ggsave(result_file(figure("japan-map"), "pdf"), width = 5, height = 5, units = "in")
```

**`r figure("japan-map")`**. `r figure_cap("japan-map")`

\clearpage

```{r raw-diversity, width = 7, height = 4}
# Richness
a <- ggplot() +
  geom_sf(data = japan, fill = "grey90", color = "black", lwd = 0.25) +
  geom_sf(data = biodiv_ferns_spatial, aes(fill = richness), color = "transparent", alpha = 0.8) +
  annotate(geom="richtext", y = Inf, x = -Inf, label = "<span style='color:black'>**A** Richness</span>", hjust = "inward", vjust = "inward", color = "white") + 
  labs(fill = "No. taxa") +
  scale_fill_scico(palette = "lajolla") + 
  scale_x_continuous(breaks = c(125, 135, 145)) +
  scale_y_continuous(breaks = c(25, 35, 45)) +
  map_theme()

@@

a + b + c +
  plot_layout(nrow = 1)

ggsave(result_file(figure("raw-diversity"), "pdf"), width = 7, height = 4, units = "in")
```

**`r figure("raw-diversity")`**. `r figure_cap("raw-diversity")` For **B** and **C**, raw branchlengths were transformed to relative values by dividing each branchlength by the sum of all branchlengths.

\clearpage

```{r rand-diversity, fig.width = 7, fig.height = 9}
# PD
a <- ggplot() +
  geom_sf(data = japan, fill = "grey90", color = "black", lwd = 0.25) +
  geom_sf(data = biodiv_ferns_spatial, aes(fill = pd_signif), color = "transparent", alpha = 0.8) +
  annotate(geom="richtext", y = Inf, x = -Inf, label = "<span style='color:black'>**A** PD</span>", hjust = "inward", vjust = "inward", color = "white") + 
  scale_fill_manual(values = signif_cols) +
  map_theme() +
  theme(
    axis.text.x = element_blank(),
    legend.title = element_blank()
    )

@@

a + b + c + d + 
  plot_layout(ncol = 2, nrow = 2, guides = 'collect') &
  theme(legend.position = "bottom")

ggsave(result_file(figure("rand-diversity"), "pdf"), width = 7, height = 8, units = "in")
```

**`r figure("rand-diversity")`**. `r figure_cap("rand-diversity")` **A**, Phylogenetic diversity (PD). **B**, Relative phylogenetic diveristy (RPD). **C**, Functional diveristy (FD). **D**, relative functional diveristy (RFD). For each metric, raw values were compared to a null distribution of 999 random values, and the *P*-value calculated using a two-tailed test. For details, see Methods.

\clearpage

```{r endemism, fig.width = 5, fig.height = 5}
ggplot() +
  geom_sf(data = japan, fill = "grey90", color = "black", lwd = 0.25) +
  geom_sf(data = biodiv_ferns_spatial, aes(fill = endem_type), color = "transparent", alpha = 0.8) +
  scale_fill_manual(values = canape_cols) +
  map_theme() +
  theme(legend.title = element_blank())

ggsave(result_file(figure("endemism"), "pdf"), width = 5, height = 5, units = "in")
```

**`r figure("endemism")`**. `r figure_cap("endemism")` 'paleo' indicates paleoendemic areas (those with an overabundance of long branches); 'neo' indicates neodendemic areas (those with an overabundance of short branches); 'mixed' indicates significantly endemic sites that have neither an overabundance of short nor long branches; 'super' indicates highly significantly endemic sites that have neither an overabundance of short nor long branches. For details, see Methods.

\clearpage

```{r bioregions, fig.width = 7, fig.height = 5}
# Taxonomic bioregions
a <- ggplot() +
  geom_sf(data = japan, fill = "grey90", color = "black", lwd = 0.25) +
  geom_sf(data = biodiv_ferns_spatial, aes(fill = taxonomic_cluster), color = "transparent", alpha = 0.8) +
  annotate(geom="richtext", y = Inf, x = -Inf, label = "<span style='color:black'>**A** Taxonomic</span>", hjust = "inward", vjust = "inward", color = "white") + 
  labs(fill = "Cluster") +
  scale_fill_manual(values = bioregion_cols) +
  map_theme()

@@

a + b +
  plot_layout(ncol = 2)

ggsave(result_file(figure("bioregions"), "pdf"), width = 7, height = 5, units = "in")
```

**`r figure("bioregions")`**. `r figure_cap("bioregions")` **A**, taxonomic bioregions. **B**, phylogenetic bioregions. Bioregions determined by clustering taxonomic (Sørensen) or phylogenetic (PhyloSor) distances into *k* = 8 or 7 groups, respectively, using the R package \package{phyloregion}.

\clearpage

```{r div-by-tax-region, fig.width = 7, fig.height = 7}
data <-
biodiv_ferns_spatial %>%
  filter(taxonomic_cluster %in% c(1,2,3,4)) %>%
  as_tibble %>%
  select(pd_obs_z, rpd_obs_z, fd_obs_z, rfd_obs_z, pe_obs_p_upper, taxonomic_cluster)

@@

a + b + c + d + e +
  plot_layout(ncol = 2) +
plot_annotation(
  tag_levels = 'A'
) &
  theme(legend.position = "none")

ggsave(result_file(figure("div-by-tax-region"), "pdf"), width = 7, height = 7, units = "in")

```

**`r figure("div-by-tax-region")`**. `r figure_cap("div-by-tax-region")` **A**, Standard effect size (SES) of phylogenetic diversity (PD). **B**, SES of relative PD (RPD). **C**, SES of functional diversity (FD). **D**, SES of relative FD (RFD). **E**, p-score of observed phylogenetic endemism (PE) relative to 999 random communities. Only top four clusters with the most sites shown.

\clearpage

```{r conserv-status, fig.width = 4, fig.height = 4}
# Plot protected areas
signif_cells_protected_area %>%
  mutate(
    metric = case_when(
      metric == "pd" ~ "PD",
      metric == "fd" ~ "FD",
      metric == "pe" ~ "PE",
      metric == "richness" ~ "Richness",
    ),
    status = str_to_sentence(status)) %>%
  mutate(metric = factor(metric, levels = c("PE", "FD", "PD", "Richness"))) %>%
  ggplot(aes(x = percent_protected, y = metric, fill = status)) +
  geom_col(position = position_dodge()) +
  geom_vline(xintercept = total_percent_protected_medium, color = "grey40", linetype = 3) +
  geom_vline(xintercept = total_percent_protected_high, color = "grey40", linetype = 2) +
  scale_fill_viridis_d() +
  scale_x_continuous(labels = scales::percent) +
  labs(
    x = "Percent protected",
    y = "Biodiversity metric",
    fill = "Protection status"
  ) +
  theme_bw() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "bottom"
  )

>>>>

---
title: "Exploring dimensions of biodiversity in Japanese ferns"
mainfont: Roboto-Regular.ttf
fontsize: 11pt
mainfontoptions: 
  - BoldFont=Roboto-Bold.ttf
  - ItalicFont=Roboto-Italic.ttf
  - BoldItalicFont=Roboto-BoldItalic.ttf
bibliography: 
  - references.yaml
  - references_other.yaml
output: 
  officedown::rdocx_document:
    base_format: bookdown::word_document2
    pandoc_args: [ "--csl", "journal-of-biogeography.csl"]
    toc: no
    reference_num: false
    reference_docx: template.docx
  bookdown::pdf_document2:
    citation_package: default
    toc: no
    number_sections: no
    fig_caption: yes
    latex_engine: xelatex
    pandoc_args: [ "--csl", "journal-of-biogeography.csl"]
params:
  doc_type: doc
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(officedown)
library(officer)
library(flextable)
library(kableExtra)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, results = "hide", cache = FALSE)

# Run these lines if rendering outside of targets plan
# library(targets)
# library(tarchetypes)
# source(here::here("R/packages.R"))
# source(here::here("R/functions.R"))
```

```{r macros}
# Define formattting for some common custom words that may vary in style between journals
ie <- "*ie*"
eg <- "*eg*"
# R packages
pack <- function(x) {glue::glue("'{x}'")}
# R functions
func <- function(x) {glue::glue("'{x}'")}
# Pagebreak for MS Word (doc) only
pagebreak_doc <- function() {ifelse(params$doc_type == "doc", run_pagebreak(), return(""))}
# Pagebreak for pdf only
pagebreak_pdf <- function() {ifelse(params$doc_type == "pdf", return("\\clearpage"), return(""))}
# Pagebreak for both
pagebreak <- function() {
  if (params$doc_type == "doc") {
    run_pagebreak()
  } else if (params$doc_type == "pdf") {
    return("\\clearpage")
  } else {
    stop("doc_type parameter must be 'doc' or 'pdf'")
  }
}
```

```{r load-data}
tar_load(c(
  "occ_point_data_ferns",
  "occ_point_data_ferns_unfiltered",
  "biodiv_ferns_spatial",
  "k_taxonomy",
  "k_phylogeny",
  "signif_cells_protected_area",
  "protected_areas",
  "japan_shp",
  "phy_sig_results",
  "binary_sig_results",
  "traits_summary",
  "fern_traits",
  "japan_pterido_tree",
  "japan_fern_tree",
  "ppgi",
  "occ_point_data_ferns",
  "comm_ferns_full",
  "comm_ferns",
  "japan_points_raw",
  # "plastome_calibration_dates",
  # FIXME: remove until phylo analysis is working "plastome_tree",
  "green_list"
))
```

@@

`r pagebreak()`

@@

We used traits originally compiled for identification of ferns and lycophytes of Japan as follows [@Ebihara2016b; @Ebihara2017]. Continuous traits were measured on 10 randomly selected specimens per species. Fertile (`r ie`, spore-bearing) and sterile fronds were measured separately, then averaged. Qualitative traits were scored by observing voucher specimens. All qualitative traits were scored in binary format; `r eg`, a single trait with three states was formatted as three binary traits. From the original trait list, we selected only traits with putative ecological function (`r table("traits-used")`), and removed any that had Pearson correlation coefficient >0.6 or fewer than three observations of a given trait state.

@@

We downloaded climate data at the 2.5 minute scale from the WorldClim database using the `r func("getData")` function in the R package `r pack("raster")`, and extracted four variables related to our research questions as follows: mean annual temperature (bio1; hereafter "temperature"), temperature seasonality (bio4), annual precipitation (bio12; hereafter "precipitation"), and precipitation seasonality (bio15). We identified the intersection of the climate data at the 2.5 minute scale with the 0.2° grid-cells, then calculated the mean of each climatic variable for each 0.2° grid-cell. The R package `r pack("sf")` was used for GIS analyses.

@@

We tested for phylogenetic signal in continuous traits using two alternative metrics, Blomberg's *K* [@Blomberg2003] and Pagel's lambda [@Pagel1999]. *K* describes the ratio between the observed variance in a trait vs. the amount of variance expected under Brownian Motion (BM); lambda is a scaling parameter that transforms the phylogeny such that trait values fit those most closely expected under BM. Both lambda and *K* = 1 under BM; lower values indicate less signal (overdispersion) and higher values indicate more signal (clustering). We measured *K* and lambda using the `r func("phylosig")` function in the R package `r pack("phytools")` [@Revell2012], which also calculates significance based on a randomization test for *K* and a likelihood ratio test for lambda. We tested for phylogenetic signal in qualitative (coded as binary) traits using Fritz and Purvis' *D* [@Fritz2010] with the `r func("phylo.d")` function in the R package `r pack("caper")` [@Orme2018], which also assesses significance by comparing the observed value to a null distribution of random values and a simulated distribution of values expected under BM.

@@

We measured phylogenetic diversity (PD) using the Faith's PD (the total branch length connecting all terminal taxa in a community, including the root of the phylogeny) [@Faith1992] with the `r func("PD")` function in the R package `r pack("phyloregion")`. To detect areas with unusually long or short branch lengths, we measured relative phylogenetic diversity (RPD), which is a ratio of observed PD to PD measured on a tree where all branch lengths have been set to equal length [@Mishler2014].

We measured functional diversity (FD) following the method of @Petchey2002. We first calculated functional distances using Gower's method [@Gower1971] on the trait data with the `r func("gowdis")` function in the R package `r pack("FD")` [@Laliberte2010]. We weighted continuous and qualitative traits equally, weighted each binary component trait of each qualitative trait equally, and log-transformed and scaled continuous traits prior to calculating distances. We then used the distances to build a trait dendrogram using the `r func("hclust")` function in the `r pack("stats")` R package [@RCoreTeam2020], and used the dendogram in place of a phylogenetic tree to calculate PD and RPD as described above. These functional analogs of PD and RPD are hereafter referred to as FD (functional diversity) and RFD (relative functional diversity).

We measured phylogenetic endemism (PE) [@Rosauer2009], which is the sum of all branches at a site inversely weighted by their range, using the `r func("phylo_endemism")` function in the R package `r pack("phyloregion")`.

@@

We analyzed bioregions using the framework of @Daru2020 and their R package `r pack("phyloregion")`. Briefly, this involves first calculating beta diversity (change in composition between communities) using either species names (taxonomic bioregions) or phylogenetic distances (phylogenetic bioregions). For taxonomic bioregions, we calculated the turnover in species between sites due to species replacement [@Baselga2012] with the Sørensen index. For phylogenetic bioregions, we calculated phylogenetic dissimilarities based on the PhyloSor index [@Bryant2008]. The distances are used to construct a dendrogram, which is then split into into *k* bioregions. As the optimal number of bioregions cannot be known *a-priori*, we tested values of *k* from 1 to 20.

@@

Grid-cells with significantly low PD were almost all located on Honshu, the main island of Japan (`r figure("rand-diversity")`A). A smaller number of cells had significantly high PD, and these were located mostly on smaller southern (Amami, Ryukyu, Izu, Ogasawara) islands or coastal areas (Izu Peninsula, Kii Peninsula). The area near Yakushima includes cells with both significantly high and low PD. Grid-cells with significantly low RPD were mostly observed in southern Honshu, particularly in and around Kyushu (`r figure("rand-diversity")`B). Grid-cells with significantly high RPD are located both in the southern islands and northern Honshu and Hokkaido. Results of the randomization test for functional diversity were broadly similar to phylogenetic diversity, but with more significant cells for FD than PD and fewer significant cells for RFD than RPD (`r figure("rand-diversity")`C, D).

@@

Nearly all grid-cells in islands south of Kyushu (Ryukyu and Ogasawara) have significant levels of phylogenetic endemism, with low (non-significant) endemism observed elsewhere (`r figure("endemism")`). The vast majority of endemic grid-cells are mixed, with very few paleo- or neoendemic cells. Small clusters of significantly endemic cells were also observed northwest of Tokyo (Nagano prefecture), and in central Hokkaido (Daisetsuzan mountains). When we ran CANAPE with a reduced dataset including only taxa endemic to Japan to account for the border effect, there were fewer significant cells, but patterns were otherwise similar to the full dataset (`r s_figure("endemism-restricted")`).

@@

The extreme evolutionary distinctiveness of southern islands is both a major feature of the Japanese fern flora and a challenge for analysis. The tropical lineages occurring in these islands are likely responsible for the comparatively low (`r ie`, clustered) values of PD observed on Honshu (`r figure("rand-diversity")`A). If a more restricted null model was used that did not consider all taxa capable of dispersing anywhere within the study area, different results would likely be obtained for the randomization test [@Mishler2020]. However, ferns are known to have high dispersal abilities [@Tryon1970; @Barrington1993; @Moran2001]. In our dataset, the taxon with the greatest latitudinal distribution, `r taxon_with_widest_lat$text`, spans nearly the entire region (`r taxon_with_widest_lat$min_lat`°N to `r taxon_with_widest_lat$max_lat`°N), and `r n_widespread_taxa` taxa have latitudinal distributions spanning the Tokara strait in the south (29°N) to the Tsugaru strait in the north (41.5°N). Therefore, it seems reasonable to assume that ferns are capable of dispersing across the entire study area.

@@

`r pagebreak()`

```{r traits-used-prep}
# Read in trait metadata
trait_signif <- read_csv(here::here("data_raw/trait_functions.csv"))

# Format traits to look nice
traits_table <- traits_summary %>%
  left_join(trait_signif, by = "trait") %>%
  mutate(
    trait = case_when(
      trait == "frond_width" ~ "Frond width",
      trait == "number_pinna_pairs" ~ "No. pinna pairs",
      trait == "stipe_length" ~ "Stipe length",
      trait == "margin" ~ "Frond margin",
      trait == "shape" ~ "Frond shape",
      trait == "texture" ~ "Frond texture",
      TRUE ~ str_to_sentence(trait)
    ),
    trait_type = case_when(
      trait_type == "qualitative" ~ glue::glue("Qualitative (coded as {n} binary traits)"),
      trait_type == "continuous" & str_detect(trait, "pinna pairs") ~ glue::glue("Continuous (integer)"),
      trait_type == "continuous" ~ glue::glue("Continuous (cm)"),
      TRUE ~ str_to_sentence(trait_type)
    )
  ) %>%
  select(trait, Significance = significance)
```

```{r traits-used-pdf, results = "asis", eval = params$doc_type == "pdf"}
# Print trait table to pdf
traits_table %>%
  column_to_rownames("trait") %>%
  knitr::kable(format = "latex", booktabs = TRUE, escape = TRUE) 
```

```{r traits-used-doc, results = "asis", eval = params$doc_type == "doc"}
# Print trait table to doc
traits_table %>%
  flextable() %>%
  set_header_labels(trait = "") %>%
  width(1, width = 1.25) %>%
  width(2, width = 4)
```

**`r table("traits-used")`**. Fern traits used in this study.

`r pagebreak()`

@@

```{r japan-map-save}
# Plot map of Japan with labels of relevant places

@@

# Make map
japan_map_plot <- ggplot(japan) + 
  geom_sf() +
  geom_richtext(
    data = japan_points, 
    aes(x = lon_adj, y = lat_adj, label = label),
    size = 2.5,
    fill = NA, label.color = NA, # remove background and outline
    label.padding = grid::unit(rep(0, 4), "pt") # remove padding)
  ) +
  geom_point(
    data = filter(japan_points, query == "Tokyo"),
    aes(x = lon, y = lat)
  ) +
  map_theme()

ggsave(result_file(figure("japan-map"), "pdf"), japan_map_plot, width = 5, height = 5, units = "in")
```

```{r japan-map-print, fig.width = 5, fig.height = 5, eval = params$doc_type == "pdf"}
# Print plot if rendering to PDF
japan_map_plot
```

**`r figure("japan-map")`**. Map of Japan showing names of places mentioned in the text. Names of the four main islands in bold.

`r pagebreak_pdf()`

```{r raw-diversity-save}
# Richness
a <- ggplot() +
  geom_sf(data = japan, fill = "grey90", color = "black", lwd = 0.25) +
  geom_sf(data = biodiv_ferns_spatial, aes(fill = richness), color = "transparent", alpha = 0.8) +
  annotate(geom="richtext", y = Inf, x = -Inf, label = "<span style='color:black'>**A** Richness</span>", hjust = "inward", vjust = "inward", color = "white") + 
  labs(fill = "No. taxa") +
  scale_fill_scico(palette = "lajolla") + 
  scale_x_continuous(breaks = c(125, 135, 145)) +
  scale_y_continuous(breaks = c(25, 35, 45)) +
  map_theme()

@@

raw_diversity_plot <- a + b + c +
  plot_layout(nrow = 1)

ggsave(result_file(figure("raw-diversity"), "pdf"), width = 7, height = 4, units = "in", raw_diversity_plot)
```

```{r raw-diversity-print, fig.width = 7, fig.height = 4, eval = params$doc_type == "pdf"}
# Print plot if rendering to PDF
raw_diversity_plot
```

**`r figure("raw-diversity")`**. Raw taxonomic (**A**), phylogenetic (**B**), and functional (**C**) diversity of the ferns of Japan. For **B** and **C**, raw branchlengths were transformed to relative values by dividing each branchlength by the sum of all branchlengths.

`r pagebreak_pdf()`

```{r rand-diversity-save}
# PD
a <- ggplot() +
  geom_sf(data = japan, fill = "grey90", color = "black", lwd = 0.25) +
  geom_sf(data = biodiv_ferns_spatial, aes(fill = pd_signif), color = "transparent", alpha = 0.8) +
  annotate(geom="richtext", y = Inf, x = -Inf, label = "<span style='color:black'>**A** PD</span>", hjust = "inward", vjust = "inward", color = "white") + 
  scale_fill_manual(values = signif_cols) +
  map_theme() +
  theme(
    axis.text.x = element_blank(),
    legend.title = element_blank()
    )

@@

rand_diversity_plot <- a + b + c + d + 
  plot_layout(ncol = 2, nrow = 2, guides = 'collect') &
  theme(legend.position = "bottom")

ggsave(result_file(figure("rand-diversity"), "pdf"), rand_diversity_plot, width = 7, height = 8, units = "in")
```

```{r rand-diversity-print, fig.width = 7, fig.height = 9, eval = params$doc_type == "pdf"}
# Print plot if rendering to PDF
rand_diversity_plot
```

**`r figure("rand-diversity")`**. Results of randomization test for phylogenetic and functional diversity of the ferns of Japan. **A**, Phylogenetic diversity (PD). **B**, Relative phylogenetic diveristy (RPD). **C**, Functional diveristy (FD). **D**, relative functional diveristy (RFD). For each metric, raw values were compared to a null distribution of 999 random values, and the *P*-value calculated using a two-tailed test. For details, see Methods.

`r pagebreak_pdf()`

```{r endemism-save, fig.width = 5, fig.height = 5}
endemism_plot <- ggplot() +
  geom_sf(data = japan, fill = "grey90", color = "black", lwd = 0.25) +
  geom_sf(data = biodiv_ferns_spatial, aes(fill = endem_type), color = "transparent", alpha = 0.8) +
  scale_fill_manual(values = canape_cols) +
  map_theme() +
  theme(legend.title = element_blank())

ggsave(result_file(figure("endemism"), "pdf"), endemism_plot, width = 5, height = 5, units = "in")
```

```{r endemism-print, fig.width = 5, fig.height = 5, eval = params$doc_type == "pdf"}
# Print plot if rendering to PDF
endemism_plot
```

**`r figure("endemism")`**. Phylogenetic endemism of the ferns of Japan measured using CANAPE (categorical analysis of neo- and paleo-endemism). 'paleo' indicates paleoendemic areas (those with an overabundance of long branches); 'neo' indicates neodendemic areas (those with an overabundance of short branches); 'mixed' indicates significantly endemic sites that have neither an overabundance of short nor long branches; 'super' indicates highly significantly endemic sites that have neither an overabundance of short nor long branches. For details, see Methods.

`r pagebreak_pdf()`

```{r bioregions-save}
# Taxonomic bioregions
a <- ggplot() +
  geom_sf(data = japan, fill = "grey90", color = "black", lwd = 0.25) +
  geom_sf(data = biodiv_ferns_spatial, aes(fill = taxonomic_cluster), color = "transparent", alpha = 0.8) +
  annotate(geom="richtext", y = Inf, x = -Inf, label = "<span style='color:black'>**A** Taxonomic</span>", hjust = "inward", vjust = "inward", color = "white") + 
  labs(fill = "Cluster") +
  scale_fill_manual(values = bioregion_cols) +
  map_theme()

@@

bioregions_plot <- a + b +
  plot_layout(ncol = 2)

ggsave(result_file(figure("bioregions"), "pdf"), bioregions_plot, width = 7, height = 5, units = "in")
```

```{r bioregions-print, fig.width = 7, fig.height = 5, eval = params$doc_type == "pdf"}
# Print plot if rendering to PDF
bioregions_plot
```

**`r figure("bioregions")`**. Bioregions of the ferns of Japan. **A**, taxonomic bioregions. **B**, phylogenetic bioregions. Bioregions determined by clustering taxonomic (Sørensen) or phylogenetic (PhyloSor) distances into *k* = 8 or 7 groups, respectively, using the R package `r pack("phyloregion")`.

`r pagebreak_pdf()`

```{r div-by-tax-region-save}
data <-
biodiv_ferns_spatial %>%
  filter(taxonomic_cluster %in% c(1,2,3,4)) %>%
  as_tibble %>%
  select(pd_obs_z, rpd_obs_z, fd_obs_z, rfd_obs_z, pe_obs_p_upper, taxonomic_cluster)

@@

div_by_tax_region_plot <- a + b + c + d + e +
  plot_layout(ncol = 2) +
plot_annotation(
  tag_levels = 'A'
) &
  theme(legend.position = "none")

ggsave(result_file(figure("div-by-tax-region"), "pdf"), div_by_tax_region_plot, width = 7, height = 7, units = "in")
```

```{r div-by-tax-region-print, fig.width = 7, fig.height = 7, eval = params$doc_type == "pdf"}
# Print plot if rendering to PDF
div_by_tax_region_plot
```

**`r figure("div-by-tax-region")`**. Phylogenetic and morphological diversity of the ferns of Japan by bioregion. **A**, Standard effect size (SES) of phylogenetic diversity (PD). **B**, SES of relative PD (RPD). **C**, SES of functional diversity (FD). **D**, SES of relative FD (RFD). **E**, p-score of observed phylogenetic endemism (PE) relative to 999 random communities. Only top four clusters with the most sites shown.

`r pagebreak_pdf()`
